---
title: "Athletes by sport type and nations"
author: "Duc-Quang Nguyen"
date: "2 Aug 2016"
output: html_document
---
## Ressources


# Icons folder needs to be copied !!

### Grouping of sports
* Grouping of olympic sports: https://en.wikipedia.org/wiki/Olympic_sports
* https://en.wikipedia.org/wiki/Combat_sport#Olympic_Combat_Sports
* Team sports: https://en.wikipedia.org/wiki/Team_sport#Olympic_team_sports
* [My own sport grouping](https://docs.google.com/spreadsheets/d/1WDkHi7tJCp2GRZPgdh4hZuSeYf5flCaeqdtnF-S4DRE/edit#gid=0)
* Summer olympics open source icons: https://commons.wikimedia.org/wiki/Category:Summer_Olympics_pictograms


## Stats

* Data last scraped: 2016-08-04 7h GMT, [google sheet](https://docs.google.com/spreadsheets/d/1WDkHi7tJCp2GRZPgdh4hZuSeYf5flCaeqdtnF-S4DRE/edit?usp=sharing)
* Swiss no teams
* Russia
  * [More than 110 Russian athletes banned](http://www.npr.org/sections/thetorch/2016/08/02/487642811/the-russian-tally-more-than-110-athletes-barred-from-the-olympics)
  * [entire weightlifting team, all but one member of the track and field team](http://www.sbnation.com/2016/8/3/12371506/119-russian-athletes-banned-rio-olympics-doping-scandal)
* Cycling
* Horse riding
* 


## Findings



```{r setup, include=FALSE}
library(readr)
library(tidyr)
library(dplyr)
library(magrittr)
library(countrycode)
library(ggplot2)
library(scales)
library(swiMap)
library(swiTheme)

### Interactive 
library(ggiraph)
library(htmltools)
library(swiRcharts)
```

```{r settings}
rio.file <- "input/athletes_rio2016.csv"
sports.file <- "input/summer Olympic sport grouping - Sheet1.csv"
trad.file <- "input/Rio althetes by sport - Sheet1.csv"
icon.file <- "input/Rio 20106 summer Olympic athletes list - sportGroup2icons.csv"


spitOutGroupSports <- F
plotDevStaticGraphic <- F
computeStats <- T

countries <- data.frame(names = c(
  "Switzerland", "Brazil", "United States", "China", "Russia", 
   "India", "Japan", "Turkey", "Germany", "Italy", #"Portugal", 
  "France", "Spain", "United Kingdom")
)
countries$ioc <- countrycode(countries$names, "country.name", "ioc")
```

```{r load data}
athletes <- read_csv(rio.file)
sports <- read_csv(sports.file)

cat("\nTotal number of athletes: " )
cat(nrow(athletes))

txt <- read.csv(trad.file, row.names = 1, stringsAsFactors = F)
# discard incomplete translations
cidx <- unique(which(txt =="" | is.na(txt), T)[,2])
if(length(cidx > 0)) {
  warning(paste(colnames(txt)[cidx], collapse = "\t"), " languages will be discarded!", "\n")
  txt <- txt[,-cidx, drop = F]
}
colnames(txt)

# check icons are present
cat("\nAssuming icons are in a 'icons/' folder")
group2icon <- read_csv(icon.file)
group2icon$icon <- paste0(ifelse(is.na(group2icon$icon), "", "icons/"), group2icon$icon)
stopifnot(all(file.exists(group2icon$icon[group2icon$icon != 'NA'])))

```

```{r wrangle data}
if(spitOutGroupSports) {
  sportGroups <- do.call(rbind, by(sports, sports$group, function(ddd) {
    data.frame(group = unique(ddd$group), sports = paste0(ddd$sport, collapse = ", "))
  }))
  write.csv(sportGroups, "data/sportGroupsForTranslation.csv")
}

idx <- match(athletes$sport, sports$sport)
stopifnot(!any(is.na(idx)))
athletes$group <- unlist(sports[idx, 'group'], use.names = F)

dat <- athletes %>% group_by(group, country, iso3) %>% summarise(value = length(athletes)) %>% ungroup()
dat %<>% group_by(country, iso3) %>% mutate(total = sum(value)) %>% ungroup()
dat$prop <- round((dat$value / dat$total), 3)

dev100 <- dat %>% group_by(country) %>% summarise(totest = sum(prop)) %>% 
  ungroup() %>% select(totest) - 1
stopifnot(abs(dev100) < .01)

data <- dat %>% filter(iso3 %in% countries$ioc)
d.all <- data %>% group_by(group) %>% 
  summarise(prop.all = (sum(value, na.rm = T) / sum(data$value))) %>% 
  ungroup() %>% arrange(desc(prop.all))
cat("\n", "Check sum overall sport proportions:", sum(d.all$prop.all))
d.all$group <- factor(d.all$group, levels = d.all$group)

# order factors
data$group <- factor(data$group, levels = d.all$group)
data$country <- gsub("Russian Federation", "Russia", data$country)
data$country <- factor(data$country, levels = as.character(countries$names))
data$iso2c <- countrycode(data$iso3, "ioc", "iso2c")

stopifnot(all(levels(data$group) %in% group2icon$group))

```


```{r visualize theme, include = F}
### ggplot2 theme
myTheme <- function(base_family = "OpenSans-CondensedLight", title_family = "OpenSans-CondensedBold") {
  swi_theme(base_size = 14, base_family = base_family, title_family = title_family) +
    theme(
      strip.text.x = element_text(family = title_family, hjust = 0, vjust = 1, size = 14),
      axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5, margin = margin(t = -17), size = 11.5),
      axis.line.x = element_blank(),
      axis.ticks.x = element_blank(),
      legend.position  = "none",
      plot.margin = unit(c(0.1, 0.3, 0, 0), "mm")
    )   
}

if(plotDevStaticGraphic) {
  ggplot(data = data, aes(x = country, y = prop, group = country, fill = country)) + 
    geom_bar(stat = "identity") + facet_wrap(~ group, ncol = 4, scales = "free_y") + 
    myTheme() + 
    scale_fill_manual(values = swi_rpal, drop=FALSE) +
    scale_x_discrete(name = "", expand = c(0,0)) +
    geom_hline(data = d.all, aes(yintercept = prop.all), linetype = 7, size = 0.2, alpha = 0.6) +
    geom_text(aes(label = iso2c), vjust=1.5, color = "#f7f5ed", family =  "OpenSans-CondensedBold", size = 3) 
}
```

## Generate each sport group as a standalone graphic

(instead of faceting)

```{r prod graphics helpers, include = F}
### print interactive chart
interactive_chart <- function(gpath, fontname = 'Open Sans Condensed') {
  ggiraph(
    code = {print(gpath)},
    hover_css = "fill-opacity:0.6;stroke-opacity:0.9;stroke-width:1.1px;stroke:#333333;",
    tooltip_opacity = 0.7,
    pointsize = 12,
    width = "100%",
    height_svg = 4.5,
    width_svg = 4,
    fontname_sans = fontname,
    fontname_serif = fontname
  ) 
}

# Plot by sport group subsets of the data (no faceting) and generate HTML text
graphGroup <- function(dd = dd, sp = sp, gr = gr, d.all = d.all, group2icon = group2icon) {
  #dd %>% filter(as.character(group) == gr)
  ddd <- dd[which(dd$group == gr),]
  
  subt1 <- txt[paste0(gr, ".underlying"), lang]  
  
  # ititle & icon
  icon.path <- group2icon[which(group2icon$group == gr), 'icon'] %>% unlist()
  icon.html <- ifelse(icon.path == 'NA', '', paste0('<img src="', icon.path, '" alt="',  gr ,' icon" height="27" width="27">'))
  tit <- paste0(icon.html,  "&nbsp; ", txt[gr, lang])
  
  ddd$tooltip <- paste0(
    "<b>", as.character(ddd$ct), "</b><br>",
    ddd$prop * 100, "%",'<div><span style="font-size: 0.8em">',
    txt[gr, lang], "<br>(", ddd$value, " / ", ddd$total, ")",
    "</span></div>"
  )    
  
  chart <- ggplot(
    data = ddd, 
    aes(
      x = ct, 
      tooltip = tooltip, 
      data_id_ = ct, 
      y = prop, 
      group = ct, 
      fill = ct)
    ) + 
    geom_bar_interactive(stat = "identity")  +
    myTheme() + 
    scale_fill_manual(values = swi_rpal, drop = F) +
    scale_y_continuous(name = "", breaks = pretty_breaks(n = 3), limits = c(0, maxV), labels = percent) + 
    scale_x_discrete(name = "", expand = c(0,0.1), drop=FALSE)
  
  gg <- chart +
    geom_hline(
      data = d.all %>% filter(group == gr), 
      aes(yintercept = prop.all), 
      linetype = 2, size = 1, alpha = 0.8, colour = "#00334d"
    ) +
    geom_text(
      aes(label = iso2c), vjust = 1.8, 
      color = "#f7f5ed", family = txt['title_family', lang], size = 3.8
    ) 
  list(chart = gg, title = tit, subtitle = subt1, pad = txt[paste0(gr,".padding"), lang])
}

maxV <- max(data$prop)

```


```{r prod graphics, include = F}

lang <- 'EN'
gr <- 'Team Sports'

for (lang in colnames(txt)) {
  dd <- data
  sp <- sports
  
  ## get tranlsations
  midx <- match(dd$iso3, row.names(txt))
  stopifnot(!any(is.na(midx)))
  dd$ct <- txt[midx, lang]
  # order country levels
  midx <- dd[match(levels(dd$country), dd$country), 'iso3'] %>% unlist()
  stopifnot(!any(is.na(midx)))
  dd$ct <- factor(dd$ct, levels = txt[match(midx, row.names(txt)), lang])
 
  midx <- match(as.character(dd$group), row.names(txt))
  stopifnot(!any(is.na(midx)))
  dd$gr <- txt[midx, lang]
  
  midx <- match(paste0(sp$group, ".underlying"), row.names(txt))
  if(any(is.na(midx))) {
    stop(paste0(sp$group, ".underlying")[is.na(midx)])
  }
  sp$gr <- txt[midx, lang]
  
  # magic happening
  charts <- lapply(levels(dd$group), function(gr) graphGroup(dd = dd, sp = sp, gr, d.all, group2icon))

  save_html(
    tags$html(
      tags$head(includeHTML("styles.html")),
      tags$body(
        h2(HTML(txt["main.title", lang])),
        div(class = "descr", HTML(paste0(txt["descr1", lang], "<br><br>", "<i>", txt["descr2", lang], "</i>"))),
        div(class = "container",
            lapply(1:length(charts), function(i) {
              div(class = "graphic", 
                  h3(HTML(charts[[i]]$title)),
                  div(class = "subtitle",  HTML(paste0(charts[[i]]$subtitle, charts[[i]]$pad))),
                  interactive_chart(charts[[i]]$chart)
              )      
            })
        ),
        div(id = "cite", HTML(paste0(txt['source', lang], ": ", txt['source.name', lang]), " | swissinfo.ch |", htmlLink("https://twitter.com/duc_qn", "@duc_qn"))),
        HTML(iframeresizer)
      )), file = paste0("rio2016_barBySportAndNations_", lang, ".html"), libdir = "js"
  )
}
```

```{r some stats}
if(computeStats) {
  
  # Countries with at least 100 athletest at Rio
  n100 <- dat %>% filter(total > 100) %>% select (country) %>% unlist %>% unique()
  length(n100)
  
  # non team
   dat %>% filter(group == 'Team Sports', total > 100) %>% 
    arrange(prop)
   team.ct100 <- dat %>% filter(group == 'Team Sports', total > 100) %>% select(country) %>% unlist()
   n100[which(!n100 %in% team.ct100)]
   
   
  # most combat nation
  dat %>% filter(group == 'Combat Sports', total > 100) %>% 
    arrange(desc(prop))

    # cycling
  dat %>% filter(group == 'Cycling', total > 100) %>% 
    arrange(desc(prop))
  dat %>% filter(group == 'Cycling') %>% 
    arrange(desc(value))

  cat("\n Out of all country delegations of at least 100 (", length(n100), "), Switzerland has the highest proportion of cylists (15.5%)") 
  
  dat %>% filter(group == 'Sailing', total > 100) %>% 
    arrange(desc(prop)) %>% head(20)
  dat %>% filter(group == 'Sailing') %>% 
    arrange(desc(value))
  
  
  dat %>% filter(group == 'Equestrian', total > 100) %>% 
    arrange(desc(prop)) %>% head(20)
  dat %>% filter(group == 'Equestrian') %>% 
    arrange(desc(value))
  
  
    
  dat %>% filter(group == 'Rowing', total > 100) %>% 
    arrange(desc(prop)) %>% head(20)
  dat %>% filter(group == 'Rowing') %>% 
    arrange(desc(value))
  
}

```